---
title: "K Nearest Neighbours"
author: "Eli Clapper"
date: "27-12-2021"
output: html_document
---

In regression problems, K-Nearest Neighbours is an algorithm where the predicted value of a case is some value that characterizes the nearest K neighbours of that case. For example, we assign the mean of the nearest three cases to the candidate case that those three points are the nearest too.

Let's illustrate in a graphical example.

Generate data with outcome and 2-dimensional input matrix
```{r}
set.seed(6164900)
Y <- rnorm(100)
X <- scale(cbind(Y + rnorm(100,0, 1.5), rnorm(100)))
df <- cbind(Y,X)

```

Plot the input data
```{r}
library(plotrix)
plot(X[,1], X[,2], col = "blue", main = "KNN with K = 3")
points(-2.1,1, col = "red")
draw.circle(-2.1,1,0.75, col = rgb(red = 1, blue = 0, green = 1, alpha = 0.3))
```
The predicted value on the outcome for the red dot depends on the three dots that closest to it. One metric that can be used is the mean on the outcome for these three blue dots. That mean will be the predicted value for the red dot. There are other metrics that can be used, such as the median, but in this example the mean is used.

There are also multiple measures that describe the 'distance' to the determine the closest neighbours. The distance between two vectors that is mostly use is the Eucledian distance, which is will be used in this example and is defined as follows:
$$
\sqrt{\sum^n_{i=1}(C_i - N_i)^2}.
$$
Where n = the number of predictors a case is observed on, C = the candidate vector and N is a neighbour vector.
We want to find the neighbours for which the root of the squared difference between vectors of the neighbour and the candidate is smallest. 

So a function that calculates Euclidean distance may look like this:
```{r}
EDist <- function(C,N){ #function to calculate Eucledian distance between vectors
  sqrt(sum((C-N)^2))
}

case <- df[1, 2:ncol(df)] #sample case (first case of df in this example)
df_wo_case <- df[-1, ] #df without the case

EDist(case, df_wo_case[1, 2:ncol(df)]) #euclidean distance of candidate and potential neighbour
```
Now that we can calculate Euclidean Distances for two cases, we can try to do it for the first case with all other cases in the data set. For 1 case, we get n-1 Euclidean Distances.

```{r}
#obtain EDs for case vs all possible neighbours
EDs1 <- sapply(1:nrow(df_wo_case), function(i){
  EDist(case, df_wo_case[i, 2:ncol(df)])
})
c(length = length(EDs1))
EDs1
```
Now we want to know the K neighbours that have the lowest Eucledian distances and take the mean over their values on the outcome variable. That would be the predicted value for the candidate. 

Now lets see make a KNN program for the first case in the data set and see how well we predict with 3 different values of K (1, 10, 80).
```{r}
predcase1 <- function(K){
  
  #obtain values of lowest K EDs
  lowest <- sort(EDs1)[1:K] 

  #index of neighbours with lowest EDs
  neighbours <- sapply(1:length(lowest), function(i){
    which(EDs1 == lowest[i])
  }) 
  
  #the outcome values for the 3 nearest neighbours
  Neighbours <- df_wo_case[neighbours, 1] 
  
  #predicted value for case 1
  pred_case1 <- mean(Neighbours) 

  return(pred_case1)
}

#plot predicted value against the actual value in the density of Y
for(i in c(1, 10, 80)){
  thispred <- predcase1(i)
  plot(density(Y), main = paste0("K = ", i))
  points(c(df[1,1], thispred), c(0.2, 0.2), col = "red")
  text(c(df[1,1]-0.4, thispred+0.55), c(0.2, 0.2), c('actual','predicted'))
}




```
We plot the predicted and actual values on the density of Y to get an idea how many standard deviations the predicted value is away from the actual. You can see that the predicted value gets better as K gets smaller. It could be argued that this is expected as it is likely that similar cases in the input data also have similar values on the outcome value. However, if we choose very small values of K, than there are very little neighbours, which makes the variance of the algorithm way bigger. A too big K however, might have little variance, but will have large bias. The best value for K is usually chosen using cross-validation. Also, with low values for K, the predicted value is based on very little data, which also could, instead of improving, worse the fit.

So now lets code a KNN for the entire dataframe and obtain predicted values and see how our function does for different values of K. We evaluate the algorithm by using certain ML metrics such as $RMSE$, $MAD$ and $R^2$.
The function requires the outcome variable to be the first column of the matrix. Afterwards we will evaluate the algorithm on a separate testing and training set.

```{r}
KNN <- function(df, K){
  
  #function to calculate Eucledian distance between vectors
  EDist <- function(C,N){ 
    sqrt(sum((C-N)^2))
  }
  
  #function that obtains KNN estimate for a single case in the df
  predict1 <- function(i){
    case <- df[i, 2:ncol(df)] #sample case (first case of df in this example)
    df_wo_case <- df[-i, ] #df without the case
    EDs <- sapply(1:nrow(df_wo_case), function(i){ #obtain
      EDist(case, df_wo_case[i, 2:ncol(df)])
    })
    lowest <- sort(EDs)[1:K] #obtain values of lowest K EDs
    
    neighbours <- sapply(1:K, function(i){
        which(EDs == lowest[i])
      }) #which neighbours have lowest EDs
      
      pred_case <- mean(df_wo_case[neighbours, 1]) #predicted value for a case
    
      return(pred_case)
  }
  
  #apply that function over all cases to obtain estimates
  y_pred <- sapply(1:nrow(df), function(x){
    predict1(x)
  })
  
  return(y_pred)
}

```

Now lets obtain metrics with the metrics function using K = 1, 10, 80, 99
```{r}
source("../Created_Functions.R")
eval <- lapply(c(1,10,80, 99), function(i){
  pred <- KNN(df, i)
  print(var(pred))
  metrics(Y, pred)
})
```
you can see that the variance of the predicted values gets lower the more cases we use as neighbours, with it being almost zero as K = nrow(df)-1. This makes sense because as we use more neighbours, the values become more similar. If K = 1, the predicted value only depends on 1 other observation. While if K = N, then every case gets assigned the mean of the entire matrix.

```{r}
names(eval) <- c("k=1", "k=10", "k=80", "k=99")
eval
```
For k = 1 and k = 99, $R^2$ is negative, indicating that just using the mean of the outcome value as a predicted value for all cases is a better option than the KNN model. In this case it seems like k = 10 does best. But it is advised to do cross-validation for more valid results.

#evaluate using cross-validation

We will know evaluate the algorithm using cross-validation. It requires a bit of recoding, because the test and train data must be put in separately. Besides, for the test data, only the input data is required to obtain estimates of the outcome.

generate data and split into train and test.
```{r}
set.seed(6164900)
Y <- rnorm(500)
X <- cbind(Y+rnorm(500, 0 , 1.2), rnorm(500))
df <- cbind(Y, X)

split <- sample(1:length(Y), size = floor(0.8*length(Y)))
train <- df[split, ]
test <- df[-split, ]
test_input <- test[, 2:ncol(test)]

```

rewrite function
```{r}
KNN_CV <- function(train, test, K){
  
  test_input <- test[,2:ncol(test)]
  ntest <- ifelse(is.null(ncol(test_input)), length(test_input), nrow(test_input)) #cases in testset
  nptest <- ifelse(is.null(ncol(test_input)), 1, 0) #predictors in testset
  
  #function to calculate Eucledian distance between vectors
  EDist <- function(C,N){ 
    sqrt(sum((C-N)^2))
  }
  
  #function that obtains KNN estimate for a single case in the df
  predict1 <- function(i){
    case <-ifelse(nptest == 1, test_input[i],test_input[i, ]) #sample case (first case of df in this example)
    EDs <- sapply(1:nrow(train), function(i){ #obtain
      EDist(case, train[i, 2:ncol(train)])
    })
    lowest <- sort(EDs)[1:K] #obtain values of lowest K EDs
    
    neighbours <- sapply(1:K, function(i){
        which(EDs == lowest[i])
      }) #which neighbours have lowest EDs
      
      pred_case <- mean(train[neighbours, 1]) #predicted value for a case
    
      return(pred_case)
  }
  
  #apply that function over all cases to obtain estimates
  y_pred <- sapply(1:ntest, function(x){
    predict1(x)
  })
  
  mets <- metrics(test[,1], y_pred)
  
  return(mets)
}
```

run function on different values of K, obtain R2. we then extract the condition for which R2 is maximum and this is the value for K we deem best.
```{r}
Ks <- seq(1, 100, 5) #conditions for K

predictions <- matrix(NA, length(Ks), 5) #create matrix to store metrics in 
colnames(predictions) <- c("RMSE", "MAD", "MaxDev", "MinDev", "R2") #give names
rownames(predictions) <- paste0("K = ", Ks)                         #give names
for(i in seq_along(Ks)){          #for every K in ks, run the CV
  predictions[i, ] <- KNN_CV(train, test, Ks[i])
} 

#obtain best condition by getting the rowname of the row where R2 is highest.
best_K <- names(which(predictions[, 'R2'] == max(predictions[, 'R2'])))
c(best_condition = best_K)
predictions[best_K, ]

```
The algorithm does okay, with an RMSE of 0.64 and R2 of 40% when K = 31.
